---
title: "Comparing sbt versions"
output:
  rmarkdown::html_vignette: default
vignette: >
  %\VignetteIndexEntry{Comparing sbt versions}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This document compares the different versions of sbt:

* v1 the status quo
* v2 introduces changes to tag likelihood
* v3 fix the multinomial likelihood for age and length composition data
* v4 removal of nuisance fisheries using cohort slicing (added as an option)
* v5 new selectivity
* v6 Dirichlet-multinomial distribution for AFs and LFs
* v7 length based M (looks similar except for tick up at end)
* v8 Estimate rec devs as random effect and estimate sigmaR
* v9 added dummy length dimension to model
* v10 state space Ns
* OSA residuals

# Load inputs

```{r echo=TRUE, message=FALSE}
library(TMB)
library(tidyverse)
library(sbt)

attach(data_csv1)

theme_set(theme_bw())
```

```{r echo=TRUE, message=FALSE}
Data <- list(last_yr = 2022, age_increase_M = 25, 
             length_m50 = 150, length_m95 = 180, 
             catch_UR_on = 0, catch_surf_case = 1, catch_LL1_case = 1, 
             scenarios_surf = scenarios_surface, scenarios_LL1 = scenarios_LL1,
             sel_min_age_f = c(2, 2, 2, 8, 6, 0), 
             sel_max_age_f = c(17, 9, 17, 22, 25, 7),
             sel_end_f = c(1, 0, 1, 1, 1, 0),
             sel_change_sd_fy = t(as.matrix(sel_change_sd[,-1])), 
             sel_smooth_sd_f = data_labrep1$sel.smooth.sd,
             pop_switch = 1, 
             hsp_switch = 1, hsp_false_negative = 0.7467647, 
             gt_switch = 1,
             cpue_switch = 1, cpue_a1 = 5, cpue_a2 = 17,
             aerial_switch = 4, aerial_tau = data_labrep1$tau.aerial, 
             troll_switch = 1, 
             af_switch = 3,
             lf_switch = 3, lf_minbin = c(1, 1, 1, 11),
             tag_switch = 1, tag_var_factor = 1.82
)

Data <- get_data(data_in = Data)
```

# Model setup

```{r get-pars, echo=TRUE, message=FALSE}
Params <- list(par_log_B0 = data_par1$ln_B0, 
               par_log_psi = log(data_par1$psi),
               par_log_m0 = log(data_par1$m0), 
               par_log_m4 = log(data_par1$m4),
               par_log_m10 = log(data_par1$m10), 
               par_log_m30 = log(data_par1$m30),
               par_log_h = log(data_par1$steep), 
               par_log_sigma_r = log(data_labrep1$sigma.r), 
               par_rdev_y = data_par1$Reps,
               par_sels_init_i = data_par1$par_sels_init_i, 
               par_sels_change_i = data_par1$par_sels_change_i,
               par_log_cpue_q = data_par1$lnq,
               par_log_cpue_sigma = log(data_par1$sigma_cpue),
               par_log_cpue_omega = log(data_par1$cpue_omega),
               par_log_aerial_tau = log(data_par1$tau_aerial),
               par_log_aerial_sel = data_par1$ln_sel_aerial,
               par_log_troll_tau = log(data_par1$tau_troll),
               par_log_hsp_q = data_par1$lnqhsp, 
               par_logit_hstar_i = qlogis(exp(data_par1$par_log_hstar_i)),
               par_log_tag_H_factor = log(data_par1$tag_H_factor),
               par_log_af_alpha = c(1, 1),
               par_log_lf_alpha = rep(1, 4)
)
```

```{r get-map, echo=TRUE, message=FALSE}
Map <- list()
Map[["par_log_psi"]] <- factor(NA)
Map[["par_log_m0"]] <- factor(NA)
Map[["par_log_m10"]] <- factor(NA)
Map[["par_log_h"]] <- factor(NA)
Map[["par_log_sigma_r"]] <- factor(NA)
Map[["par_log_cpue_sigma"]] <- factor(NA)
Map[["par_log_cpue_omega"]] <- factor(NA)
Map[["par_log_aerial_tau"]] <- factor(NA)
Map[["par_log_aerial_sel"]] <- factor(rep(NA, 2))
Map[["par_log_troll_tau"]] <- factor(NA)
Map[["par_log_hsp_q"]] <- factor(NA)
Map[["par_log_tag_H_factor"]] <- factor(NA)
```

```{r compile-models, echo=TRUE, message=FALSE}
sbt_src <- "../sbt/src"

compile(file.path(sbt_src, "sbt_v1.cpp"))
compile(file.path(sbt_src, "sbt_v2.cpp"))
# compile(file.path(sbt_src, "sbt_v3.cpp"))

dyn.load(dynlib(file.path(sbt_src, "sbt_v1")))
dyn.load(dynlib(file.path(sbt_src, "sbt_v2")))
# dyn.load(dynlib(file.path(sbt_src, "sbt_v3")))
```

# Status quo

```{r model-1, echo=TRUE, message=FALSE}
obj1 <- MakeADFun(data = Data, parameters = Params, map = Map, DLL = "sbt_v1")

bnd <- get_bounds(obj = obj1)

opt1 <- nlminb(start = obj1$par, objective = obj1$fn, gradient = obj1$gr, 
               lower = bnd$lower, upper = bnd$upper,
               control = list(eval.max = 20000, iter.max = 20000))
```

# Changes to tag likelihood

```{r model-2, echo=TRUE, message=FALSE}
Map2 <- Map
Map2[["par_logit_hstar_i"]] <- NULL
Map2[["par_log_af_alpha"]] <- factor(rep(NA, 2))
Map2[["par_log_lf_alpha"]] <- factor(rep(NA, 4))

obj2 <- MakeADFun(data = Data, parameters = Params, map = Map2, DLL = "sbt_v2")

opt2 <- nlminb(start = obj2$par, objective = obj2$fn, gradient = obj2$gr, 
               lower = bnd$lower, upper = bnd$upper,
               control = list(eval.max = 20000, iter.max = 20000))
```

# Use proper multinomial distribution

```{r model-3, echo=TRUE, message=FALSE}
Data3 <- Data
Data3$af_switch <- 0
Data3$lf_switch <- 0

obj3 <- MakeADFun(data = Data3, parameters = Params, map = Map2, DLL = "sbt_v2")

opt3 <- nlminb(start = obj3$par, objective = obj3$fn, gradient = obj3$gr, 
               lower = bnd$lower, upper = bnd$upper,
               control = list(eval.max = 20000, iter.max = 20000))
```

# Try Dirichlet-multinomial

```{r model-4, echo=TRUE, message=FALSE}
Data4 <- Data
Data4$af_switch <- 2
Data4$lf_switch <- 2

Map4 <- Map
Map4[["par_log_af_alpha"]] <- NULL
Map4[["par_log_lf_alpha"]] <- NULL

obj4 <- MakeADFun(data = Data4, parameters = Params, map = Map4, DLL = "sbt_v2")

bnd4 <- get_bounds(obj = obj4)
bnd4 %>% filter(parameter %in% c("par_log_af_alpha", "par_log_lf_alpha"))

opt4 <- nlminb(start = obj4$par, objective = obj4$fn, gradient = obj4$gr, 
               lower = bnd4$lower, upper = bnd4$upper,
               control = list(eval.max = 20000, iter.max = 20000))

exp(opt4$par[grepl("par_log_af_alpha", names(opt4$par))])
exp(opt4$par[grepl("par_log_lf_alpha", names(opt4$par))])

hist(obj4$report()$lp_af)
hist(obj4$report()$lp_lf)
hist(obj4$report()$lp_tags)
which(obj4$report()$lp_tags == max(obj4$report()$lp_tags))
```

# Try Dirichlet

```{r model-5, echo=TRUE, message=FALSE}
Data5 <- Data
Data5$af_switch <- 1
Data5$lf_switch <- 1

Params5 <- Params
Params5$par_logit_hstar_i <- NULL

Map5 <- Map
Map5[["par_log_af_alpha"]] <- NULL
Map5[["par_log_lf_alpha"]] <- NULL

obj5 <- MakeADFun(data = Data5, parameters = Params5, map = Map5, DLL = "sbt_v2")
obj5$fn(obj5$par)
obj5$report()$lp_af
obj5$report()$lp_lf
Data$lf_n[is.nan(obj5$report()$lp_lf)]
Data$lf_n
Data$af_min_age
obs <- (Data$af_obs[1,] / sum(Data$af_obs[1,])) + 1e-6
obs / sum(obs)

bnd5 <- get_bounds(obj = obj5)
bnd5 %>% filter(parameter %in% c("par_log_af_alpha", "par_log_lf_alpha"))

opt5 <- nlminb(start = obj5$par, objective = obj5$fn, gradient = obj5$gr, 
               lower = bnd5$lower, upper = bnd5$upper,
               control = list(eval.max = 20000, iter.max = 20000))

exp(opt5$par[grepl("par_log_af_alpha", names(opt5$par))])
exp(opt5$par[grepl("par_log_lf_alpha", names(opt5$par))])
```


```{r model-5, echo=TRUE, message=FALSE}
Map3 <- Map
Map3[["par_log_sigma_r"]] <- NULL

Map4 <- Map
Map4[["par_log_sigma_r"]] <- NULL
Map4[["par_log_m0"]] <- NULL
Map4[["par_log_m10"]] <- NULL

# obj3 <- MakeADFun(data = Data, parameters = Params, map = Map3, 
#                   random = "par_rdev_y", DLL = "sbt_v2")
# obj4 <- MakeADFun(data = Data, parameters = Params, map = Map4, 
#                   random = "par_rdev_y", DLL = "sbt_v2")
# obj5 <- MakeADFun(data = Data, parameters = Params, map = Map, 
#                   DLL = "sbt_v3")

# opt4 <- nlminb(start = obj4$par, objective = obj4$fn, gradient = obj4$gr, 
#                lower = bnd$lower, upper = bnd$upper)
# opt5 <- nlminb(start = obj5$par, objective = obj5$fn, gradient = obj5$gr, 
#                lower = bnd$lower, upper = bnd$upper)
# 
# exp(Params$par_log_sigma_r)
# exp(opt3$par["par_log_sigma_r"])

```

# Compare models

```{r echo=FALSE, message=FALSE}
table_compare <- function(xlist) {
  n <- length(xlist)
  pars1 <- c("R0", "par_B0", "par_psi", 
             "par_log_cpue_q", "par_log_cpue_sigma", "par_log_cpue_omega", 
             "par_log_aerial_tau", "par_log_troll_tau", "par_hsp_q", "par_log_tag_H_factor", 
             "par_m0", "par_m4", "par_m10", "par_m30", 
             "par_log_af_alpha", "par_log_lf_alpha",
             "par_sigma_r", "par_h", "alpha", "beta", "tau_ac2")
  dl1 <- list()
  for (i in 1:n) {
    rp <- xlist[[i]]$report()
    vals <- unlist(rp[which(names(rp) %in% pars1)])
    dl1[[i]] <- data.frame(Model = i, Output = names(vals), value = as.numeric(vals))
  }
  pars2 <- c("lp_cpue", "lp_aerial", "lp_aerial_tau",
             "lp_troll", "lp_tags", "lp_pop", "lp_hsp", "lp_gt", 
             "lp_af", "lp_lf", "lp_rec", "lp_sel", "nll")
  dl2 <- list()
  for (i in 1:n) {
    rp <- xlist[[i]]$report()
    vlist <- rp[which(names(rp) %in% pars2)]
    vals <- unlist(lapply(vlist, sum))
    dl2[[i]] <- data.frame(Model = i, Output = names(vals), value = as.numeric(vals))
  }
  df <- bind_rows(dl1, dl2) %>%
    pivot_wider(names_from = Model) %>%
    mutate(Output = factor(Output, levels = c(pars1, pars2))) %>%
    arrange(Output)
  return(df)
}

tb <- table_compare(xlist = list(obj1, obj2, obj3, obj4, obj5))

print(tb, n = 100)

# library(kableExtra)
# 
# df = tb[,-1]
# fn <- function(x) cell_spec(x, bold = T, color = spec_color(x, end = 0.9))    
# df[, 1:2] = t(apply(df[,1:2], 1, fn))  
# 
# df %>%
#   kable(escape = F, align = "c") %>%
#   kable_styling(c("striped", "condensed"), full_width = F)

# kbl(tb, digits = Inf, caption = NULL) %>%
#   kable_styling("bordered")# %>%
  # row_spec(row = 2, color = "white", background = spec_color(as.numeric(tb[2,]))) %>%
  # row_spec(row = 4, color = "white", background = spec_color(as.numeric(tb[4,])))
```

```{r plot-compare-optim, echo=FALSE, message=FALSE, fig.cap="Spawning biomass (tonnes)."}
plot_compare <- function(x) {
  n <- length(x)
  
  dl <- list()
  for (i in 1:n) {
    dl[[i]] <- data.frame(Year = 1931:2023, 
                          SSB = x[[i]]$report()$spawning_biomass_y / 1e6,
                          Model = paste0("v", i))
  }
  df <- bind_rows(dl)
  
  ggplot(data = df, aes(x = Year, y = SSB, colour = Model, linetype = Model)) +
    geom_line() + 
    scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
    labs(x = "Year", y = "Spawning biomass (millions of tonnes)")
}

plot_compare(x = list(obj1, obj2, obj3, obj4))

plot_compare(x = list(obj2, obj5))

plot_compare(x = list(obj1, obj2, obj3, obj4, obj5))

obj <- obj5
plot_cpue(data = Data, object = obj)
# plot_sliced_afs(data = Data)
plot_af(data = Data, object = obj, fishery = "Indonesian")
plot_af(data = Data, object = obj, fishery = "Australian")
plot_lf(data = Data, object = obj, fishery = "LL1")
plot_lf(data = Data, object = obj, fishery = "LL2")
plot_lf(data = Data, object = obj, fishery = "LL3")
plot_lf(data = Data, object = obj, fishery = "LL4")
```

# OSA residuals

```{r osa-resids, echo=FALSE, message=FALSE, fig.cap="OSA residuals."}
library(compResidual)

# Age comps ----

X <- t(Data$af_obs * Data$af_n)
P <- t(obj1$report()$af_pred)

colSums(X)
colSums(P)

Data$af_min_age
Data$af_max_age

X5 <- X[7:31, Data$af_fishery == 5]
P5 <- P[7:31, Data$af_fishery == 5]
min(X5); min(P5)
res5 <- resMulti(obs = X5, pred = P5)
colnames(res5) <- c(Data$first_yr:Data$last_yr)[Data$af_year[Data$af_fishery == 5]]
plot(res5)
plot(res5, pick_one = 1)

X6 <- X[1:8, Data$af_fishery == 6]
P6 <- P[1:8, Data$af_fishery == 6]
min(X6); min(P6)
res6 <- resMulti(obs = X6 + 1e-6, pred = P6)
colnames(res6) <- c(Data$first_yr:Data$last_yr)[Data$af_year[Data$af_fishery == 6]]
plot(res6)

# Length comps ----

X <- t(Data$lf_obs * Data$lf_n)
P <- t(obj1$report()$lf_pred)

i <- 1
Xi <- X[,Data$lf_fishery == i]
Pi <- P[,Data$lf_fishery == i]
min(Xi); min(Pi)
resi <- resMulti(obs = Xi, pred = Pi)
colnames(resi) <- c(Data$first_yr:Data$last_yr)[Data$lf_year[Data$lf_fishery == i]]
plot(resi)

# plot(res, pick_one = 1)
# plot(res, pick_one = 3)
# plot(res, pick_one = 4)
# res <- resDirM(obs = X, alpha = P + 1e-6)
```
