---
title: "OMMP16"
format:
  html:
    embed-resources: true
    toc: true
    toc-location: left
    lightbox: true
    theme: cosmo
    code-fold: true
    code-tools: true
editor_options: 
  chunk_output_type: console
---

# Load inputs

The `sbt` RTMB model is loaded along with several R functions using 
`library(sbt)`. The `tidyverse` and `reshape2` packages are used for data 
manipulation and plotting. The code `theme_set(theme_bw())` alters the plot 
aesthetics of the entire article:

```{r}
#| label: load-pkg
# remotes::install_github("janoleko/RTMBdist")
# remotes::install_github("andrjohns/StanEstimators")
# remotes::install_github("noaa-afsc/SparseNUTS")

library(tidyverse)
library(sbt)
library(SparseNUTS)
library(bayesplot)

theme_set(theme_bw())

data_loc <- "csv_2026"

length_mean <- read_csv(file.path(data_loc, "mean_length.csv"))
length_sd <- read_csv(file.path(data_loc, "sd_length.csv"))
catch <- read_csv(file.path(data_loc, "catch.csv"))
catch_UA <- read_csv(file.path(data_loc, "catch_UA.csv"))
scenarios_surface <- read_csv(file.path(data_loc, "scenarios_surface.csv"))
scenarios_LL1 <- read_csv(file.path(data_loc, "scenarios_LL1.csv"))
POPs <- read_csv(file.path(data_loc, "POPs.csv"))
HSPs <- read_csv(file.path(data_loc, "HSPs.csv"))
GTs <- read_csv(file.path(data_loc, "GTs.csv"))
troll <- read_csv(file.path(data_loc, "trolling_index.csv"))
cpue <- read_csv(file.path(data_loc, "cpue.csv"))
age_freq <- read_csv(file.path(data_loc, "age_freq.csv"))
length_freq <- read_csv(file.path(data_loc, "lf_assessment.csv")) # this is the new version from Jim with the CPUE LFs

data <- list(
  last_yr = 2025, age_increase_M = 25,
  length_m50 = 150, length_m95 = 180, 
  catch_UR_on = 0, catch_surf_case = 1, catch_LL1_case = 1, 
  scenarios_surf = scenarios_surface, scenarios_LL1 = scenarios_LL1,
  removal_switch_f = c(0, 0, 0, 1, 0, 0), # 0=harvest rate, 1=direct removals
  # removal_switch_f = c(0, 0, 0, 0, 0, 0), # 0=harvest rate, 1=direct removals
  sel_min_age_f = c(2, 2, 2, 8, 6, 0, 2),
  # sel_max_age_f = c(17, 9, 17, 22, 25, 7, 17),
  sel_max_age_f = c(17, 9, 17, 21, 25, 7, 17),
  sel_end_f = c(1, 0, 1, 1, 1, 0, 1), # 0=zero, 1=constant
  sel_LL1_yrs = c(1952, 1957, 1961, 1965, 1969, 1973, 1977, 1981, 1985, 1989, 1993, 1997, 2001, 2006, 2007, 2008, 2011, 2014, 2017, 2020, 2023),
  sel_LL2_yrs = c(1969, 2001, 2005, 2008, 2011, 2014, 2017, 2020),
  sel_LL3_yrs = c(1954, 1961, 1965, 1969, 1970, 1971, 2005, 2006, 2007),
  sel_LL4_yrs = c(1953),
  sel_Ind_yrs = c(1976, 1995, 1997, 1999, 2002, 2004, 2006, 2008, 2010, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023),
  sel_Aus_yrs = c(1952, 1969, 1973, 1977, 1981, 1985, 1989, 1993, 1997, 1998, 1999, 2000, 2001, 2002, 2003, 2004, 2005, 2006, 2007, 2008, 2009, 2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023),
  sel_CPUE_yrs = c(1969, 1973, 1977, 1981, 1985, 1989, 1993, 1997, 2001, 2006, 2007, 2008, 2011, 2014, 2017, 2020),
  # af_switch = 9,
  # af_switch = 1, # CAUSES ISSUES WHY?
  af_switch = 2, # 1=multinomial, 2=Dirichlet, 3=Dirichlet-multinomial, 9=old
  lf_switch = 2, lf_minbin = c(1, 1, 1, 11),
  cpue_switch = 1, cpue_a1 = 5, cpue_a2 = 17,
  aerial_switch = 4, aerial_tau = 0.3, 
  troll_switch = 1, 
  pop_switch = 1, 
  hsp_switch = 1, 
  hsp_false_negative = 0.6840729, # If I set the hsp false negative to 1 and af_switch = 9 then model OK
  gt_switch = 1,
  tag_switch = 1, tag_var_factor = 1.82
)

data <- get_data(data_in = data)

# Need to fix LFs ----

repair_af_slices_weighted <- function(catch_ysf,
                                      af_ysfa,
                                      N = 3L,
                                      p = 1,
                                      eps = 1e-12,
                                      expand_if_empty = TRUE,
                                      verbose = TRUE) {
  # af_ysfa is assumed [year, season, fishery, age]
  d_af <- dim(af_ysfa)
  stopifnot(length(d_af) == 4)
  n_y <- d_af[1]; n_s <- d_af[2]; n_f <- d_af[3]; n_a <- d_af[4]
  
  d_c <- dim(catch_ysf)
  stopifnot(length(d_c) == 3)
  
  # map catch dims to y/s/f by matching sizes in af
  map_dim <- function(target, used) {
    idx <- which(d_c == target & !(seq_along(d_c) %in% used))
    if (length(idx) != 1) return(NA_integer_)
    idx
  }
  used <- integer(0)
  iy <- map_dim(n_y, used); used <- c(used, iy)
  is <- map_dim(n_s, used); used <- c(used, is)
  iff <- map_dim(n_f, used)
  
  if (anyNA(c(iy, is, iff))) {
    stop(sprintf(
      "Could not align catch_ysf dims %s to af_ysfa y/s/f dims %s. ",
      paste(d_c, collapse = "x"),
      paste(c(n_y, n_s, n_f), collapse = "x")
    ))
  }
  
  # helper to read catch regardless of dim order
  get_catch <- function(y, s, f) {
    idx <- vector("list", 3)
    idx[[iy]] <- y; idx[[is]] <- s; idx[[iff]] <- f
    do.call(`[`, c(list(catch_ysf), idx, list(drop = TRUE)))
  }
  
  repaired <- vector("list", 0L)
  
  for (f in seq_len(n_f)) {
    for (s in seq_len(n_s)) {
      af_sum_y <- vapply(seq_len(n_y), function(yi) sum(af_ysfa[yi, s, f, ], na.rm = TRUE), numeric(1))
      usable_y <- which(af_sum_y > eps)
      
      for (y in seq_len(n_y)) {
        cval <- get_catch(y, s, f)
        asum <- af_sum_y[y]
        
        if (!is.na(cval) && cval > 0 && (is.na(asum) || asum <= eps)) {
          lo <- max(1L, y - N); hi <- min(n_y, y + N)
          cand <- setdiff(seq.int(lo, hi), y)
          donors <- intersect(cand, usable_y)
          if (length(donors) == 0L && expand_if_empty) donors <- setdiff(usable_y, y)
          if (length(donors) == 0L) {
            stop(sprintf("Cannot repair AF at y=%d s=%d f=%d: no donor years with nonzero AF.", y, s, f))
          }
          
          dist <- abs(donors - y); dist[dist == 0] <- 1e-9
          w <- (1 / (dist^p)); w <- w / sum(w)
          
          donor_mat <- vapply(donors, function(yy) af_ysfa[yy, s, f, ], numeric(n_a))
          af_new <- as.numeric(donor_mat %*% w)
          af_new[is.na(af_new)] <- 0
          z <- sum(af_new)
          if (z <= eps) stop(sprintf("Repair produced zero AF at y=%d s=%d f=%d.", y, s, f))
          af_ysfa[y, s, f, ] <- af_new / z
          af_sum_y[y] <- 1.0
          
          repaired[[length(repaired) + 1L]] <- data.frame(
            year = y, season = s, fishery = f, n_donors = length(donors)
          )
        }
      }
    }
  }
  
  repaired_df <- if (length(repaired)) do.call(rbind, repaired) else data.frame()
  if (verbose && nrow(repaired_df) > 0) {
    message(sprintf("Repaired %d year-season-fishery AF slice(s).", nrow(repaired_df)))
  }
  list(af_ysfa = af_ysfa, repaired = repaired_df, catch_dim_map = c(year = iy, season = is, fishery = iff))
}

for (f in seq_len(data$n_fishery)) {
  if (data$removal_switch_f[f] > 0) {
    x1 <- data$catch_obs_ysf[,,f]
    x2 <- apply(data$af_sliced_ysfa[,,f,], 1:2, sum)
    # x1 <- proj_catch_ysf[,,f]
    # x2 <- apply(proj_af_sliced_ysfa[,,f,], 1:2, sum)
    df1 <- as.data.frame(x1) %>%
      rownames_to_column("Year") %>%
      pivot_longer(cols = c(`1`, `2`), names_to = "Season", values_to = "catch")
    df2 <- as.data.frame(x2) %>%
      rownames_to_column("Year") %>%
      pivot_longer(cols = c(`1`, `2`), names_to = "Season", values_to = "AF")
    df <- left_join(df1, df2, by = c("Year", "Season")) %>%
      mutate(Year = as.numeric(Year)) %>%
      mutate(violation = (catch > 0 & AF == 0) | (AF > 0 & catch == 0)) %>%
      filter(violation)
    if (nrow(df) > 0) warning("You have specified catch for a fishery using direct removals but no associated sliced AFs.")
    # data$catch_obs_ysf[39:44,1,f]
    # rowSums(data$af_sliced_ysfa[60:65,1,f,])
  }
}
df

# obs_fix <- repair_af_slices_weighted(
#   catch_ysf = data$catch_obs_ysf,
#   af_ysfa   = data$af_sliced_ysfa,
#   N = 3, p = 1
# )
# data$af_sliced_ysfa <- obs_fix$af_ysfa

# proj_fix <- repair_af_slices_weighted(
#   catch_ysf = proj_catch_ysf,
#   af_ysfa   = proj_af_sliced_ysfa,
#   N = 3, p = 1
# )
# proj_af_sliced_ysfa <- proj_fix$af_ysfa
```

# Model setup

Define the parameter `list`:

```{r}
#| label: get-pars
parameters <- get_parameters(data = data)
exp(parameters$par_log_h)
parameters$par_log_h <- log(0.8)
```

There is a lot of flexibility in specifying priors now. The default priors are 
loaded using `get_priors` and the `evaluate_priors` function can be used to 
check that the priors are doing what you expect:

```{r}
#| label: get-priors
data$priors <- get_priors(parameters = parameters)
evaluate_priors(parameters = parameters, priors = data$priors)
```

```{r}
#| echo: false
#| results: asis
cat(priors_to_math(data$priors), sep = "\n")
```

Use RTMB's `map` option to turn parameters on/off:

```{r}
#| label: get-map
map <- get_map(parameters = parameters)
# map$par_log_af_alpha <- NULL
# map$par_log_lf_alpha <- factor(c(1, 2, 3, NA, 4))
```

Using the `data`, the `parameters`, the parameter `map`, and the model 
(`sbt_model`), the AD object is created using RTMBs `MakeADFun` function:

```{r}
#| label: run-nlminb
#| cache: true
obj <- MakeADFun(func = cmb(sbt_model, data), 
                 parameters = parameters, map = map)
unique(names(obj$par))
obj$fn(obj$par)
bounds <- get_bounds(obj = obj, parameters = parameters)

control <- list(eval.max = 10000, iter.max = 10000)

opt <- nlminb(start = obj$par, objective = obj$fn, gradient = obj$gr, hessian = obj$he,
              lower = bounds$lower, upper = bounds$upper, control = control)
opt <- nlminb(start = opt$par, objective = obj$fn, gradient = obj$gr, hessian = obj$he,
              lower = bounds$lower, upper = bounds$upper, control = control)

invisible(obj$fn(opt$par))
obj$env$last.par.best[1:13]
```

Check that all parameters are estimable using the `check_estimability` function.
This function was taken from the `TMBhelper` package and added to the `sbt` 
package because the `TMBhelper` package is not set up as a proper package (it 
does not install properly on the GitHub servers).

```{r}
#| label: check-estimability
#| echo: true
#| message: true
#| warning: false
check_estimability(obj = obj)
# ce <- check_estimability(obj = obj)
# ce[[4]] %>% filter(Param_check != "OK")
```

# Model fits

Model fit to CPUE (@fig-cpue-fit) and the aerial surveys (@fig-aerial-fit).
Length frequency fits for each longline fleet are shown in @fig-plot-LL1,
@fig-plot-LL2, @fig-plot-LL3, and @fig-plot-LL4.
Age composition fits are shown for the Indonesian fleet (@fig-plot-Indo) and
the Australian fleet (@fig-plot-Aus).

```{r}
#| label: fig-cpue-fit
#| echo: true
#| message: false
#| fig-cap: Model fits to CPUE.
plot_cpue(data = data, object = obj, nsim = 10)
```

```{r}
#| label: fig-aerial-fit
#| echo: true
#| message: false
#| fig-cap: Model fits to aerial surveys.
plot_aerial_survey(data = data, object = obj, nsim = 10)
```

```{r}
#| label: fig-plot-LL1
#| echo: true
#| message: false
#| fig-height: 14
#| fig-cap: Model fits to LL1 length frequencies.
plot_lf(data = data, object = obj, fishery = "LL1", ncol = 5)
```

```{r}
#| label: fig-plot-LL2
#| echo: true
#| message: false
#| fig-height: 8
#| fig-cap: Model fits to LL2 length frequencies.
plot_lf(data = data, object = obj, fishery = "LL2", ncol = 5)
```

```{r}
#| label: fig-plot-LL3
#| echo: true
#| message: false
#| fig-height: 14
#| fig-cap: Model fits to LL3 length frequencies.
plot_lf(data = data, object = obj, fishery = "LL3", ncol = 5)
```

```{r}
#| label: fig-plot-LL4
#| echo: true
#| message: false
#| fig-height: 9
#| fig-cap: Model fits to LL4 length frequencies.
plot_lf(data = data, object = obj, fishery = "LL4", ncol = 5)
```

```{r}
#| label: fig-plot-Indo
#| echo: true
#| message: false
#| fig-height: 8
#| fig-cap: Model fits to Indonesian age compositions.
plot_af(data = data, object = obj, fishery = "Indonesian")
```

```{r}
#| label: fig-plot-Aus
#| echo: true
#| message: false
#| fig-height: 10
#| fig-cap: Model fits to Australian age compositions.
plot_af(data = data, object = obj, fishery = "Australian")
```

I think that not all of the catch is being taken for LL4 because there are years 
where there is catch but no LFs but we've set this fishery to direct removals 
(@fig-catch-fit, @fig-catch-resid).

```{r}
#| label: fig-catch-fit
#| echo: true
#| message: false
#| fig-cap: Catch.
plot_catch(data = data, object = obj)
```

```{r}
#| label: fig-catch-resid
#| echo: true
#| message: false
#| fig-cap: Catch residuals.
plot_catch(data = data, object = obj, plot_resid = TRUE)
```

# One step ahead (OSA) residuals

One step ahead (OSA) residuals are a replacement for Pearson residuals 
(@fig-osa-resid).

```{r}
#| label: fig-osa-resid
#| echo: true
#| message: false
#| fig-cap: One step ahead (OSA) residuals.
plot_cpue_residuals(data = data, obj = obj, type = "OSA")
```

## Derived quantities

Recruitment deviates and recruitment (@fig-recruitment-dev, @fig-recruitment).
Natural mortality at age is shown in @fig-M.
Initial numbers at age are shown in @fig-init-n.
Harvest rates are shown in @fig-hrate.

```{r}
#| echo: false
get_link <- function(df) {
  csv_text <- capture.output(write.csv(df, row.names = FALSE))
  csv_encoded <- base64enc::base64encode(charToRaw(paste(csv_text, collapse = "\n")))
  sprintf(
    '<a href="data:text/csv;base64,%s" download="data.csv" style="text-decoration: underline; color: blue;">download csv</a>',
    csv_encoded
  )
}

download_link <- data.frame(Year = data$first_yr:(data$last_yr + 1), Recruitment = obj$report()$number_ysa[,1,1]) %>% 
  get_link()
```

```{r}
#| label: fig-recruitment-dev
#| echo: true
#| message: false
#| fig-cap: !expr paste0("Recruitment deviates (", download_link, ").")
plot_rec_devs(data = data, object = obj)
```

```{r}
#| label: fig-recruitment
#| echo: true
#| message: false
#| fig-cap: !expr paste0("Recruitment (", download_link, ").")
plot_recruitment(data = data, object = obj)
```

```{r}
#| label: fig-M
#| echo: true
#| message: false
#| fig-cap: Natural mortality at age.
plot_natural_mortality(data = data, object = obj)
```

```{r}
#| label: fig-init-n
#| echo: true
#| message: false
#| fig-cap: Initial numbers at age.
plot_initial_numbers(data = data, object = obj)
```

```{r}
#| label: fig-hrate
#| echo: true
#| message: false
#| fig-cap: Harvest rate by year.
plot_hrate(data = data, object = obj, years = 1990:2010)
```

In the plots below, the different coloured blocks of years indicate blocks of 
parameters, vertical dashed lines indicate the minimum and maximum ages over 
which selectivity is estimated, and crosses to the left indicate that LF data is 
available during that year.
Selectivity at age by year is shown for the LL1 (@fig-plot-sel-ll1),
LL2 (@fig-plot-sel-ll2), LL3 (@fig-plot-sel-ll3), Indonesian (@fig-plot-sel-indo),
Australian (@fig-plot-sel-aus), and CPUE (@fig-plot-sel-cpue) fleets.
Spawning biomass over time is shown in @fig-plot-sbio.

```{r}
#| label: fig-plot-sel-ll1
#| echo: true
#| message: false
#| fig-height: 10
#| fig-cap: Selectivity at age by year for the LL1 fleet.
# yrs <- c(1952, 1953, 1954, seq(1955, 2020, 5), 2021, 2022)
yrs <- c(1952:2022)
plot_selectivity(data = data, object = obj, years = yrs, fisheries = "LL1")
```

```{r}
#| label: fig-plot-sel-ll2
#| echo: true
#| message: false
#| fig-height: 10
#| fig-cap: Selectivity at age by year for the LL2 fleet.
plot_selectivity(data = data, object = obj, years = yrs, fisheries = "LL2")
```

```{r}
#| label: fig-plot-sel-ll3
#| echo: true
#| message: false
#| fig-height: 10
#| fig-cap: Selectivity at age by year for the LL3 fleet.
plot_selectivity(data = data, object = obj, years = yrs, fisheries = "LL3")
```

```{r}
#| label: fig-plot-sel-ll4
#| echo: true
#| message: false
#| fig-height: 10
#| fig-cap: Selectivity at age by year for the LL4 fleet.
# plot_selectivity(data = data, object = obj, years = yrs, fisheries = "LL4")
```

```{r}
#| label: fig-plot-sel-indo
#| echo: true
#| message: false
#| fig-height: 10
#| fig-cap: Selectivity at age by year for the Indonesian fleet.
plot_selectivity(data = data, object = obj, years = yrs, fisheries = "Indonesian")
```

```{r}
#| label: fig-plot-sel-aus
#| echo: true
#| message: false
#| fig-height: 10
#| fig-cap: Selectivity at age by year for the Australian fleet.
plot_selectivity(data = data, object = obj, years = yrs, fisheries = "Australian")
```

```{r}
#| label: fig-plot-sel-cpue
#| echo: true
#| message: false
#| fig-height: 10
#| fig-cap: Selectivity at age by year for the CPUE index.
plot_selectivity(data = data, object = obj, years = yrs, fisheries = "CPUE")
```

```{r}
#| label: fig-plot-sbio
#| echo: true
#| message: false
#| fig-cap: !expr paste0("Spawning biomass by year (", download_link, ").")
plot_biomass_spawning(data_list = list(data), object_list = list(obj))
```
