---
title: "Comparison with the ADMB model"
format:
  html:
    toc: true
    lightbox: true
vignette: >
  %\VignetteIndexEntry{Comparison with the ADMB model}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
editor_options: 
  chunk_output_type: console
---
  
```{r}
#| include: false
library(knitr)
opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
#| label: data-and-params
#| echo: false
#| message: false
#| results: hide
library(reshape2)
library(tidyverse)
library(sbt)
library(kableExtra)
library(testthat)

theme_set(theme_bw())

attach(data_csv1)
lr <- data_labrep1

data <- list(
  last_yr = 2022, age_increase_M = 25,
  length_m50 = 150, length_m95 = 180, 
  catch_UR_on = 0, catch_surf_case = 1, catch_LL1_case = 1, 
  scenarios_surf = scenarios_surface, scenarios_LL1 = scenarios_LL1,
  sel_min_age_f = c(2, 2, 2, 8, 6, 0), 
  sel_max_age_f = c(17, 9, 17, 22, 25, 7),
  sel_end_f = c(1, 0, 1, 1, 1, 0),
  sel_change_sd_fy = t(as.matrix(sel_change_sd[,-1])), 
  sel_smooth_sd_f = lr$sel.smooth.sd,
  pop_switch = 1, 
  hsp_switch = 1, hsp_false_negative = 0.7467647, 
  gt_switch = 1,
  cpue_switch = 1, cpue_a1 = 5, cpue_a2 = 17,
  aerial_switch = 4, aerial_tau = data_labrep1$tau.aerial, 
  troll_switch = 0, 
  lf_minbin = c(1, 1, 1, 11),
  tag_switch = 1, tag_var_factor = 1.82
)

data <- get_data_v1(data_in = data)

parameters <- list(
  par_log_B0 = data_par1$ln_B0, par_log_psi = log(data_par1$psi),
  par_log_m0 = log(data_par1$m0), par_log_m4 = log(data_par1$m4),
  par_log_m10 = log(data_par1$m10), par_log_m30 = log(data_par1$m30),
  par_log_h = log(data_par1$steep), par_log_sigma_r = log(lr$sigma.r), 
  par_log_cpue_q = data_par1$lnq,
  par_log_cpue_sigma = log(data_par1$sigma_cpue),
  par_log_cpue_omega = log(data_par1$cpue_omega),
  par_log_aerial_tau = log(data_par1$tau_aerial),
  par_log_aerial_sel = data_par1$ln_sel_aerial,
  par_log_troll_tau = log(data_par1$tau_troll),
  par_log_hsp_q = data_par1$lnqhsp, 
  par_logit_hstar_i = qlogis(exp(data_par1$par_log_hstar_i)),
  par_log_tag_H_factor = log(data_par1$tag_H_factor),
  par_rdev_y = data_par1$Reps,
  par_sels_init_i = data_par1$par_sels_init_i, 
  par_sels_change_i = data_par1$par_sels_change_i
)

get_bounds_v1 <- function(obj, parameters) {
  Lwr <- rep(-Inf, length(obj$par))
  Upr <- rep(Inf, length(obj$par))
  Lwr[grep("par_log_m0", names(obj$par))] <- log(0.2) # These were the old M bounds
  Upr[grep("par_log_m0", names(obj$par))] <- log(0.55)
  Lwr[grep("par_log_m4", names(obj$par))] <- parameters$par_log_m10
  Upr[grep("par_log_m4", names(obj$par))] <- log(0.333 * exp(parameters$par_log_m10) + 0.667 * exp(parameters$par_log_m0))
  Lwr[grep("par_log_m10", names(obj$par))] <- log(0.029)
  Upr[grep("par_log_m10", names(obj$par))] <- log(0.21)
  Lwr[grep("par_log_m30", names(obj$par))] <- log(0.2)
  Upr[grep("par_log_m30", names(obj$par))] <- log(0.7)
  Upr[grep("par_log_troll_tau", names(obj$par))] <- log(0.4)
  Lwr[grep("par_rdev_y", names(obj$par))] <- rep(-5, length(parameters$par_rdev_y))
  Upr[grep("par_rdev_y", names(obj$par))] <- rep(5, length(parameters$par_rdev_y))
  df <- data.frame(parameter = names(obj$par), init = obj$par, lower = Lwr, upper = Upr)
  return(df)
}
```

```{r}
#| label: get-map
#| echo: false
#| message: false
map <- list()
map[["par_log_psi"]] <- factor(NA)
map[["par_log_m0"]] <- factor(NA)
map[["par_log_m10"]] <- factor(NA)
map[["par_log_h"]] <- factor(NA)
map[["par_log_sigma_r"]] <- factor(NA)
map[["par_log_cpue_sigma"]] <- factor(NA)
map[["par_log_cpue_omega"]] <- factor(NA)
map[["par_log_aerial_tau"]] <- factor(NA)
map[["par_log_aerial_sel"]] <- factor(rep(NA, 2))
map[["par_log_hsp_q"]] <- factor(NA)
map[["par_log_tag_H_factor"]] <- factor(NA)
```

```{r}
#| label: build-obj
#| echo: false
#| message: false
#| results: hide
obj <- MakeADFun(func = cmb(sbt_model_v1, data), parameters = parameters, map = map)
```

# Introduction

This vignette compares outputs from the Automatic Differentiation Model Builder 
(ADMB) operating model (OM) to the `RTMB` OM within the 
`sbt` package. This is done using a [fixed parameter](#fixed) run and then by 
[optimising](#optim). In the fixed parameter run the parameter values from a single cell 
of the ADMB model are used as fixed value inputs to `sbt` and outputs from `sbt` 
are generated then compared to the ADMB outputs.

# Fixed parameter run {#fixed}

## Comparison table

@tbl-comparison includes transformed parameters (i.e., in natural space), 
derived quantities, priors, 
penalties, likelihoods, and objective function values for the ADMB model, the RTMB 
model, and the difference and percent difference between the two models.

```{r}
#| label: tbl-comparison
#| echo: false
#| message: false
rp <- obj$report()

Label <- c("B0", "psi", "sigmaR", "h", "q HSP", "m0", "m4", "m10", "m30", "R0", "alpha", "beta", "tau_ac2")

df0 <- data.frame(
  Label = Label,
  ADMB = c(lr$B0, lr$psi, lr$sigma.r, lr$steep, lr$qhsp, lr$M[c(1, 5, 11, 31)], lr$R0[1], lr$alpha[1], lr$beta[1], lr$AC_penalty[2]),
  TMB = c(rp$par_B0, rp$par_psi, rp$par_sigma_r, rp$par_h, rp$par_hsp_q, rp$par_m0, rp$par_m4, rp$par_m10, rp$par_m30, rp$R0, rp$alpha, rp$beta, rp$tau_ac2)) %>%
  mutate(diff1 = ADMB - TMB) %>%
  mutate(diff2 = abs(diff1 / ADMB) * 100) %>%
  mutate(diff2 = ifelse(is.finite(diff2), diff2, 0))

Label <- names(lr$penal)
Label[13] <- "sel.init"

df1 <- data.frame(
  Label = Label,
  ADMB = lr$penal,
  TMB = c(0, rp$lp_sel[1:2], rp$lp_rec, 0, 0, 0, 0, 0, 0, 0, 0, rp$lp_sel[3], rp$lp_hstar)) %>%
  mutate(diff1 = ADMB - TMB) %>%
  mutate(diff2 = abs(diff1 / ADMB) * 100) %>%
  mutate(diff2 = ifelse(is.finite(diff2), diff2, 0)) %>%
  filter(Label != "NA")

lp_lf <- data.frame(lp = rp$lp_lf, fishery = data$lf_fishery) %>%
  group_by(fishery) %>%
  summarise(lp = sum(lp))

lp_af <- data.frame(lp = rp$lp_af, fishery = data$af_fishery) %>%
  group_by(fishery) %>%
  summarise(lp = sum(lp))

df2 <- data.frame(
  Label = c(names(lr$lnlike), "ObjF"),
  ADMB = c(as.numeric(lr$lnlike), lr$ObjF),
  TMB = c(lp_lf$lp, lp_af$lp, sum(rp$lp_cpue), sum(rp$lp_tags), sum(rp$lp_aerial) + rp$lp_aerial_tau, sum(rp$lp_troll), sum(rp$lp_pop), sum(rp$lp_hsp), sum(rp$lp_gt), rp$nll)) %>%
  mutate(diff1 = ADMB - TMB) %>%
  mutate(diff2 = abs(diff1 / ADMB) * 100) %>%
  mutate(diff2 = ifelse(is.finite(diff2), diff2, 0))

df <- bind_rows(df0, df1, df2)

kbl(
  x = df, 
  col.names = c("", "ADMB", "RTMB", "Difference", "Percent difference"), 
  caption = "Comparison of ADMB and RTMB model outputs including transformed parameters, derived quantities, priors, penalties, likelihoods, and objective function values.", 
  digits = c(7, 7, 7, 7, 4)
) %>%
  kable_styling("bordered") %>%
  column_spec(column = 5, color = "white", background = spec_color(df$diff2)) %>%
  group_rows("Parameters", 1, 9) %>%
  group_rows("Derived quantities", 10, 13) %>%
  group_rows("Priors & penalties", 14, 24) %>%
  group_rows("Likelihoods & objective function", 25, nrow(df))
```

## Comparison figures

The following figures compare outputs from the ADMB and RTMB models:
average weight at age (@fig-plot-weight), natural mortality at age (@fig-plot-M),
initial numbers at age (@fig-plot-init), recruitment deviates (@fig-plot-rec-devs),
recruitment (@fig-plot-recs), total numbers (@fig-plot-N),
spawning biomass (@fig-plot-sbio), CPUE (@fig-plot-cpue),
phi at age (@fig-phi-age), selectivity at age (@fig-sel-compare),
and age-frequency observations and predictions (@fig-af-obs, @fig-af-pred).

```{r}
#| label: define-function
#| echo: false
#| message: false
plot_compare <- function(x, ADMB, sbt, xlab, ylab) {
  data.frame(x = x, ADMB = ADMB, sbt = sbt) %>%
    pivot_longer(cols = ADMB:sbt, names_to = "Model") %>%
    ggplot(aes(x = x, y = value, colour = Model, linetype = Model)) +
    geom_line() + 
    labs(x = xlab, y = ylab) +
    scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05)))
}
```

```{r}
#| label: fig-plot-weight
#| echo: false
#| message: false
#| fig-cap: Average weight (kg) at age.
plot_compare(x = 0:30, ADMB = lr$spweights.age[1,], sbt = data$weight_fya[3,1,], 
             xlab = "Age", ylab = "Weight (kg)")
```

```{r}
#| label: fig-plot-M
#| echo: false
#| message: false
#| fig-cap: Natural mortality (M) at age.
plot_compare(x = 0:30, ADMB = lr$M, sbt = rp$M_a, xlab = "Age", ylab = "Natural mortality (M)")
```

```{r}
#| label: fig-plot-init
#| echo: false
#| message: false
#| fig-cap: Initial numbers at age in the population.
num <- rp$number_ysa
plot_compare(x = 0:30, ADMB = lr$Ns[1,] / 1e6, sbt = num[1,1,] / 1e6, xlab = "Age", ylab = "Initial numbers (millions)")
```

```{r}
#| label: fig-plot-rec-devs
#| echo: false
#| message: false
#| fig-cap: Recruitment deviates each year.
plot_compare(x = 1931:2022, ADMB = lr$Rdev, sbt = rp$rdev_y, xlab = "Year", ylab = "Recruitment deviates") +
  scale_y_continuous(limits = c(NA, NA))
```

```{r}
#| label: fig-plot-recs
#| echo: false
#| message: false
#| fig-cap: Recruitment each year.
plot_compare(x = 1931:2023, ADMB = lr$Recruitment / 1e6, 
             sbt = rp$recruitment_y / 1e6,
             # sbt = num[,1,1] / 1e6,
             xlab = "Year", ylab = "Recruitment (million)")
```

```{r}
#| label: fig-plot-N
#| echo: false
#| message: false
#| fig-cap: Total number of inidividuals each year.
plot_compare(x = 1931:2023, ADMB = rowSums(lr$Ns), sbt = rowSums(num[,1,]), xlab = "Year", ylab = "Total number")
```

```{r}
#| label: fig-plot-sbio
#| echo: false
#| message: false
#| fig-cap: Spawning biomass (tonnes).
plot_compare(x = 1931:2023, ADMB = lr$Sbio / 1e6, sbt = rp$spawning_biomass_y / 1e6, 
             xlab = "Year", ylab = "Spawning biomass (millions of tonnes)")
```

```{r}
#| label: fig-plot-cpue
#| echo: false
#| message: false
#| fig-cap: Catch per unit effort (CPUE).
plot_compare(x = data$cpue_years + data$first_yr, ADMB = lr$cpue.pred, sbt = rp$cpue_pred, xlab = "Year", ylab = "CPUE")
```

```{r}
#| label: fig-phi-age
#| echo: false
#| message: false
#| fig-cap: Phi at age for a subset of years.
yrs <- c(1952, seq(1960, 2020, 10), 2022)

x1 <- melt(lr$phi) %>%
  mutate(year = c(1931:2022)[Var2], Age = Var1 - 1) %>%
  filter(year %in% yrs)

x2 <- melt(rp$phi_ya) %>%
  mutate(year = c(1931:2022)[Var1], Age = Var2 - 1) %>%
  filter(year %in% yrs)

ggplot(data = x1, aes(x = Age, y = value)) +
  geom_line() +
  geom_line(data = x2, color = "red", linetype = "dashed") +
  facet_wrap(year ~ .) +
  labs(y = "Phi") +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05)))
```

```{r}
#| label: fig-sel-compare
#| echo: false
#| message: false
#| fig-height: 10
#| fig-cap: Selectivity at age for a subset of years.
yrs <- c(1952, seq(1960, 2020, 10), 2022)

x1 <- data.frame(lr$sel) %>%
  rename(fishery = X1, year = X2) %>%
  pivot_longer(cols = starts_with("X"), names_to = "Age") %>%
  mutate(Age = as.numeric(gsub("X", "", Age)) - 3) %>%
  filter(year %in% yrs)

x2 <- rp$sel_fya %>%
  melt() %>%
  rename(fishery = Var1) %>%
  mutate(year = c(1931:2022)[Var2], Age = Var3 - 1) %>%
  filter(year %in% yrs)

ggplot(data = x1, aes(x = Age, y = value)) +
  geom_line() +
  geom_line(data = x2, color = "red", linetype = "dashed") +
  facet_grid(year ~ fishery) +
  labs(y = "Selectivity") +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05)))
```

There was a reason why the LF observations are different 
(@fig-lf-obs) but I cannot remember why. Must be something to do 
with post processing in the code as the likelihoods are the same.

```{r}
#| label: fig-lf-obs
#| echo: false
#| message: false
#| fig-height: 10
#| fig-cap: LF observations for a subset of years.
yrs <- c(1952, seq(1960, 2020, 10), 2022)

df1 <- data.frame(data_labrep1$len.obs) %>%
  rename(fishery = X1, year = X2) %>%
  pivot_longer(starts_with("X")) %>% 
  mutate(length = parse_number(name) - 2)

oo <- data$lf_obs
colnames(oo) <- NULL
df2 <- data.frame(fishery = data$lf_fishery, year = data$lf_year + data$first_yr, oo) %>%
  pivot_longer(starts_with("X")) %>% 
  mutate(length = parse_number(name))

ggplot(data = df1 %>% filter(year %in% yrs), aes(x = length, y = value)) +
  geom_line() +
  geom_line(data = df2 %>% filter(year %in% yrs), colour = "red", linetype = "dashed") +
  facet_grid(year ~ fishery) +
  labs(x = "Length (cm)") +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05)))
```

```{r}
#| label: fig-af-obs
#| echo: false
#| message: false
#| fig-height: 10
#| fig-cap: AF observations for a subset of years.
yrs <- c(1964, seq(1995, 2020, 5), 2022)

df5 <- data.frame(data_labrep1$age.obs.1) %>%
  rename(fishery = X1, year = X2) %>%
  pivot_longer(starts_with("X")) %>% 
  mutate(age = parse_number(name) - 3 + 6)
df6 <- data.frame(data_labrep1$age.obs.2) %>%
  rename(fishery = X1, year = X2) %>%
  pivot_longer(starts_with("X")) %>% 
  mutate(age = parse_number(name) - 3)
df1 <- rbind(df5, df6) %>% select(-name)

io <- data$af_obs
colnames(io) <- NULL
df2 <- data.frame(fishery = data$af_fishery, year = data$af_year + data$first_yr, io) %>%
  pivot_longer(starts_with("X")) %>% 
  mutate(age = parse_number(name) - 1) %>% 
  select(-name)

dfa <- full_join(df1, df2, by = join_by(fishery, year, age)) %>%
  relocate(fishery, year, age) %>%
  arrange(fishery, year, age) %>%
  filter(!is.na(value.x))
# sum((dfa$value.x - dfa$value.y)^2)

ggplot(data = df1 %>% filter(year %in% yrs), aes(x = age, y = value)) +
  geom_line() +
  geom_line(data = df2 %>% filter(year %in% yrs), colour = "red", linetype = "dashed") +
  facet_grid(year ~ fishery) +
  labs(x = "Age") +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05)))
```

```{r}
#| label: fig-af-pred
#| echo: false
#| message: false
#| fig-cap: AF predictions for a subset of years.
#| fig-height: 10
yrs <- c(1964, seq(1995, 2020, 5), 2022)

df5 <- data.frame(data_labrep1$age.pred_1) %>%
  rename(fishery = X1, year = X2) %>%
  pivot_longer(starts_with("X")) %>% 
  mutate(age = parse_number(name) - 3 + 6)
df6 <- data.frame(data_labrep1$age.pred_2) %>%
  rename(fishery = X1, year = X2) %>%
  pivot_longer(starts_with("X")) %>% 
  mutate(age = parse_number(name) - 3)
df1 <- rbind(df5, df6) %>% select(-name)

df2 <- data.frame(fishery = data$af_fishery, year = data$af_year + data$first_yr, obj$report()$af_pred) %>%
  pivot_longer(starts_with("X")) %>% 
  mutate(age = parse_number(name) - 1) %>% 
  select(-name)

dfa <- full_join(df1, df2, by = join_by(fishery, year, age)) %>%
  relocate(fishery, year, age) %>%
  arrange(fishery, year, age) %>%
  filter(!is.na(value.x))

# range(dfa$value.x - dfa$value.y)

ggplot(data = df1 %>% filter(year %in% yrs), aes(x = age, y = value)) +
  geom_line() +
  geom_line(data = df2 %>% filter(year %in% yrs), colour = "red", linetype = "dashed") +
  facet_grid(year ~ fishery) +
  labs(x = "Age") +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05)))
```

# Optimising {#optim}

```{r}
#| label: run-nlminb
#| echo: false
#| results: hide
bnd <- get_bounds_v1(obj, parameters)
control <- list(eval.max = 10000, iter.max = 10000)
# opt <- nlminb(start = obj$par, objective = obj$fn, gradient = obj$gr, hessian = obj$he,
#               lower = bnd$lower, upper = bnd$upper, control = control)
# opt <- nlminb(start = opt$par, objective = obj$fn, gradient = obj$gr, hessian = obj$he,
#               lower = bnd$lower, upper = bnd$upper, control = control)
# save(obj, file = "inst/extdata/admb.rda")
load(system.file("extdata", "admb.rda", package = "sbt"))
```

The `sbt` model was optimised using `nlminb` to see if `RTMB` produces the same result.
The `nlminb` function finds the same optimum as ADMB (@fig-plot-compare-optim).

```{r}
#| label: fig-plot-compare-optim
#| echo: false
#| message: false
#| fig-cap: Spawning biomass (tonnes).
rp <- obj$report()
# sdr <- sdreport(obj = obj, getJointPrecision = TRUE)
# he <- obj$he()
# he_inv <- solve(he)
# he_inv[1:3, 1:3]
# c2 <- cov2cor(he_inv)
# range(sqrt(diag(he_inv)))
# ii <- which(sqrt(diag(he_inv)) > 21)
# names(obj$par)[ii]
# he_inv[ii]
# he_ch <- chol(he)
# ev <- eigen(he)
# range(ev$values)
# ce <- check_estimability(obj = obj)
# ce[[4]] %>% filter(Param_check != "OK")
# obj$par[1:10]
# obj$report()$spawning_biomass_y
# opt$par[1:10]
# obj$par <- opt$par
plot_compare(x = 1931:2023, ADMB = lr$Sbio / 1e6, sbt = rp$spawning_biomass_y / 1e6, 
             xlab = "Year", ylab = "Spawning biomass (millions of tonnes)")
```

# Grid {#grid}

Checking the grid.

```{r}
#| label: fig-grid
#| echo: false
#| message: false
#| fig-cap: Grid.
run_grid_v1 <- function(data, grid_parameters, bounds, map = list(), random = c(), control = list(eval.max = 10000, iter.max = 10000), parallel = FALSE, n_passes = 3) {
  n_grid <- length(grid_parameters)
  grid_obj <- vector("list", length = n_grid)
  for (i in seq_len(n_grid)) {
    obj <- MakeADFun(func = cmb(sbt_model_v1, data), parameters = grid_parameters[[i]], map = map, random = random)
    opt <- NULL
    start <- obj$par
    for (pass in seq_len(n_passes)) {
      opt <- nlminb(start = start, objective = obj$fn, gradient = obj$gr, hessian = obj$he, lower = bounds$lower, upper = bounds$upper, control = control)
      start <- opt$par
      if (opt$convergence == 0) break
    }
    obj$par <- opt$par
    obj$env$last.par.best <- opt$par
    obj$fn(opt$par)
    obj$opt <- opt
    grid_obj[[i]] <- obj
    message("Done: ", i, " of ", n_grid, ". Converged: ", opt$convergence == 0, ". Code: ", opt$convergence, ". Objective: ", opt$objective)
  }
  names(grid_obj) <- names(grid_parameters)
  return(grid_obj)
}

sample_grid_v1 <- function(grid, n_samples = 2000, prior_h = NULL, prior_psi = c(0.25, 0.5, 0.25), seed = NULL) {
  n_grid <- length(grid)
  gdf <- matrix(NA, nrow = n_grid, ncol = 5)
  colnames(gdf) <- c("nll", "m0", "m10", "h", "psi")
  pb <- txtProgressBar(min = 1, max = n_grid, style = 3)
  
  for (i in 1:n_grid) {
    rep <- grid[[i]]$report()
    gdf[i, 1] <- grid[[i]]$fn()
    gdf[i, 2] <- rep$par_m0
    gdf[i, 3] <- rep$par_m10
    gdf[i, 4] <- rep$par_h
    gdf[i, 5] <- rep$par_psi
    setTxtProgressBar(pb, i)
  }
  
  gdf <- data.frame(gdf) %>% 
    mutate(ll = -(nll - min(nll)))
  
  # Create default uniform priors if none are provided
  unique_h <- unique(gdf$h)
  if (is.null(prior_h)) {
    prior_h <- rep(1 / length(unique_h), length(unique_h))
  }
  df_prior_h <- tibble(h = unique_h, w_h = as.numeric(prior_h))
  unique_psi <- unique(gdf$psi)
  if (is.null(prior_psi)) {
    prior_psi <- rep(1 / length(unique_psi), length(unique_psi))
  }
  df_prior_psi <- tibble(psi = unique_psi, w_psi = as.numeric(prior_psi))
  
  # Calculate cell probabilities using both group likelihoods and prior weights
  df <- gdf %>%
    summarise(sum_exp_ll = sum(exp(ll)), .by = c(h, psi)) %>%
    full_join(gdf, by = join_by(h, psi)) %>%
    left_join(df_prior_h, by = join_by(h)) %>%
    left_join(df_prior_psi, by = join_by(psi)) %>%
    mutate(prob = w_h * w_psi * (exp(ll) / sum_exp_ll))
  
  if (!is.null(seed)) {
    set.seed(seed)
  }
  grid_ints <- sample(x = 1:n_grid, size = n_samples, replace = TRUE, prob = df$prob)
  
  grid_freq <- gdf %>% 
    mutate(grid_ints = as.factor(1:n())) %>% 
    left_join(data.frame(table(grid_ints)), by = join_by(grid_ints)) %>%
    mutate(Freq = ifelse(is.na(Freq), 0, Freq)) %>%
    mutate(Cell = as.integer(grid_ints)) %>%
    relocate(Cell) %>%
    select(-grid_ints) %>%
    tibble()
  
  return(list(grid_cells = grid_ints, grid_freq = grid_freq))
}

grid_to_tmbfit_v1 <- function(data, parameters, grid, grid_parameters, grid_cells) {
  # map <- get_map(parameters = parameters)
  map <- list()
  map[["par_log_psi"]] <- factor(NA)
  map[["par_log_m0"]] <- factor(NA)
  map[["par_log_m10"]] <- factor(NA)
  map[["par_log_h"]] <- factor(NA)
  map[["par_log_sigma_r"]] <- factor(NA)
  map[["par_log_cpue_sigma"]] <- factor(NA)
  map[["par_log_cpue_omega"]] <- factor(NA)
  map[["par_log_aerial_tau"]] <- factor(NA)
  map[["par_log_aerial_sel"]] <- factor(rep(NA, 2))
  map[["par_log_hsp_q"]] <- factor(NA)
  map[["par_log_tag_H_factor"]] <- factor(NA)
  map$par_log_psi <- NULL
  map$par_log_m0 <- NULL
  map$par_log_m10 <- NULL
  map$par_log_h <- NULL
  obj <- MakeADFun(func = cmb(sbt_model_v1, data), parameters = parameters, map = map)
  parnames <- names(obj$par)
  n_samples <- length(grid_cells$grid_cells)
  samples <- array(dim = c(n_samples, 1, length(obj$par)), dimnames = list(NULL, NULL, parnames))
  for (i in 1:n_samples) {
    j <- grid_cells$grid_cells[i]
    # obj$env$parList() alternatively could use this below?
    samples[i, 1,] <- c(grid[[j]]$par[1], # par_log_B0
                        grid_parameters[[j]]$par_log_psi, 
                        grid_parameters[[j]]$par_log_m0, 
                        grid[[j]]$par[2], # par_log_m4
                        grid_parameters[[j]]$par_log_m10,
                        grid[[j]]$par[3], # par_log_m30
                        grid_parameters[[j]]$par_log_h,
                        grid[[j]]$par[-c(1:3)])
  }
  # getS3method("print", "tmbfit")
  # Create tmbfit object
  x <- list(samples = samples, sampler_params = list(), mle = NULL,
            monitor = NULL, model = "RTMB", metric = "", par_names = parnames, 
            max_treedepth = NA, warmup = 0, iter = n_samples, thin = 1, 
            time.warmup = 0, time.sampling = NA, time.total = NA, algorithm = "grid")
  class(x) <- c("tmbfit", "list")
  return(x)
}

grid_pars <- get_grid(parameters = parameters, h = c(0.63, 0.72), psi = c(1.75))
grid_list <- run_grid_v1(data = data, grid_parameters = grid_pars, bounds = bnd, map = map, control = control, parallel = FALSE)
# # grid_check <- check_grid(grid = grid_list)
grid_cells <- sample_grid_v1(grid = grid_list, seed = 42, prior_psi = c(1))
grid_tmbfit <- grid_to_tmbfit_v1(data = data, parameters = parameters, grid = grid_list, grid_parameters = grid_pars, grid_cells = grid_cells)
save(grid_tmbfit, file = "inst/extdata/grid_old.rda")

load(system.file("extdata", "grid_old.rda", package = "sbt"))
barplot(table(exp(grid_tmbfit$samples[,,'par_log_m0'])), xlab = 'm0')
barplot(table(factor(exp(grid_tmbfit$samples[,,'par_log_m10']), levels = c(0.065, 0.085, 0.105))), xlab = 'm10')
```
